\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-18 writing/manuscript.Rmd
cleaned text and added refs

<!--
@ciuti and @berger-tal: human presence in nature for reasons other than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
-->

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_20129; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs. From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$). As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns as well as infrequent and unpredictable events. Examples of predictable changes in $\var(R)$ include changes that occur between seasons, such as changes in temperature in continental summers (e.g., between 15 and 30 $^\circ$C) and winter temperatures (e.g., from -40 to -10$^\circ$C within the span of a few days), changes in precipitation frequency and unpredictability between the dry season (predictable lack of rain) and the wet season (frequent but sudden and unpredictable rains), or changes in fruit and berry abundance with the arrival of spring and summer (since the timing between plants can depend on many unaccounted factors). Although each of these events is unpredictable, the change in unpredictability can be predicted and expected. In contrast, other sources of stochasticity can be much less predictable, such as fires, floods, droughts, the arrival of new competitors, and changes in climate [@ipcc_global_2018; @noonan_situ_2018].

@@

The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs [@boratynski_energetic_2020; @noonan_effects_2020], so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality. A subsequent analysis by @nilsen_can_2005 found that various carnivores respond to resource variability and seasonality (albeit in contrasting ways) using the Canadian lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*). Additionally, @nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$, so it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two. It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants. Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels.

@@

\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019]. However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\var(R)$ will have an appreciable effect on the location's favourableness. Since $\var(R)$ is not constant over time, we can now define it as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b). In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).

Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years. In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples]. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as during pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one [@lucherini_habitat_1996].

Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how $\vpos$ remains constant. Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant). Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g. from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$. In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures). However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time. Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

@@

The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant. Overall, $\vpos$ increases with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the 50% core home range ($H_{50\%}$). ***here*** <!--Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).-->

Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the core home range ($H_{50\%}$). Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios. Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes. Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The relationships were estimated using a Generalized Linear Model with a Gamma family. The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile. Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

@@

#### Move the following paragraph to previous sections

@nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region. However, the effect of many other factors on HR size likely depends on the species, diet type [@harestad_home_1979], location, and possibly also on the individual animal. For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; Capuchin monkeys, @torrez-herrera_monkeys_2020]. Similarly, predation may force animals to move more frequently to escape predators, or it may prevent them from venturing too far from the safety of their core HR too often [@suraci_beyond_2022]. Patch quality, size, fragmentation, and heterogeneity may cause animals to explore more patches if some are of low value, too small, too disconnected, or too variable [@fahrig_is_2019], but high diversity may also decrease HR size if animals require heterogeneous habitats [@fox_niche_1981; @lucherini_habitat_1996]. Similarly, patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006]. However, not all animals take advantage of linear features or higher predictability in human-altered habitats. @noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed. Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat. Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.

@@

Simulation refs: @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021

* failing to account for $\var(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
* find examples of extreme events that caused population collapses or mass die-offs
* increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness
* animal mortality following fires, how little we know of mortality following fires: @jolly_animal_2022

@@

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months. @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring. However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@abrahms_memory_2019]. @merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.

@@

In the case of animals whose spatial needs change over their lifetimes [@refs], one may define a HR as a function of time with distinct HRs for each period [salmon: @ref; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017] or as a continuous function of time. In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available. However, when an animal's spatial needs change over time, models which assume range residency are not appropriate. Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results. Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].

@@

# Extra notes

@polansky_elucidating_2015 as example of memory and changes in movement type (exploration vs stationary)

@teitelbaum_how_2015 found that resource abundance (measured as NDVI) had a strong negative effect on migration distance, while the variance in NDVI only had a small positive effect on **???**.

high-resource areas and explaining 23% of the variation in migration distances.

>>>>

<!--
@ciuti and @berger-tal: human presence in nature for reasons other than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
-->

@@

\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref]. The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022]. Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)). As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].

Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. While competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].

It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose: @singh_migration_2012; eagles: @wheat_migrate_2017; @poessel_interpreting_2022; flamingos: @pretorius_movement_2020]. From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$). As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.

\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019] although it will not be possible to predict where high-$R$ locations are, since $\e(R)$ does not vary predictably over space.

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns as well as infrequent and unpredictable events. Examples of predictable changes in $\var(R)$ include changes that occur between seasons, such as changes in temperature in continental summers (e.g., between 15 and 30 $^\circ$C) and winter temperatures (e.g., from -40 to -10$^\circ$C within the span of a few days), changes in precipitation frequency and unpredictability between the dry season (predictable lack of rain) and the wet season (frequent but sudden and unpredictable rains), or changes in fruit and berry abundance with the arrival of spring and summer (since the timing between plants can depend on many unaccounted factors). Although each of these events is unpredictable, the change in unpredictability can be predicted and expected. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors, fires, floods, droughts, and changes in climate [@jolly_animal_2022; @ipcc_global_2018; @noonan_situ_2018].

@@

When the (un)predictability in $R$ chances over time (figure \@ref(fig:multipanel-changing-var)a), changes in $\var(R)$ may have an appreciable effect on the location's favourableness. In this scenario, we can define $\var(R)$ as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b). When both the expected $R$ and its unpredictability change over time, accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).

Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years. In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs. These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022]. Thus, failing to account for changes in $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].

The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs [@boratynski_energetic_2020; @noonan_effects_2020], so the degree to which $R$ varies will likely affects each species differently. Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not detect such differences. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic. In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions. Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality. A subsequent analysis by @nilsen_can_2005 found that various carnivores respond to resource variability and seasonality (albeit in contrasting ways) using the Canadian lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*). Additionally, @nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$, so it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two. It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants. Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels [@harestad_home_1979; @noonan_effects_2020].

@@

In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable. Thus, we expect $\var(R)$ to have a stronger effect on resource abundance when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.

Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$. The top row (constant $\var(R)$) shows how simulated spatial needs vary for different trends in $\e(R)$ while $var(R)$ remains constant. As $\e(R)$ increases at a constant slope (linear $\e(R)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home range size approaches the minimum size required by the animal. This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g., from 1 to 2) correspond to large multiplicative changes (i.e., 2 is twice as big as 1). Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ because larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the home range.

@@

In regions where the average resource abundance changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures). However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable decreases in $\e(R)$ (such as following fires) may increase animals' spatial needs unpredictably. Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible. In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\e(R)$).

The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated home range while $\e(R)$ remains constant. Overall, both quantiles increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$, as with changes in $\e(R)$. Additionally, note how both the estimated home range size and the noise in the estimate increase with $\var(R)$ (e.g., constant $\e(R)$, linear $\var(R)$). This is because as the habitat becomes more stochastic, the amount of space required to reach satiety also becomes increasingly unpredictable.

The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be. Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both change monotonically, especially when the trends are linear. When both $\e(R)$ and $\var(R)$ increase linearly, the change in $H_{95\%}$ is small an not monotonic, while there is no visible change in $H_{50\%}$. However, when $\e(R)$ changes nonlinearly, it is easy to see how the increase in $\var(R)$ amplifies the effects of $\e(R)$. Similarly, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$). Thus, the estimated spatial needs are largest when $\e(R)$ is low and $\var(R)$ is high and they are smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$, and @lai_movement_2017].

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios. $\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ are relatively small and remain undetected. Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease). Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict. Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if changes are stochastic but sufficiently small and frequent, animals may perceive them as few smooth changes rather a series of small, sudden, changes. Estimating trends in $\e(R)$ and $\var(R)$ requires a high cognitive capacity, since distinguishing between $\var (R)$ and changes in $\e(R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent (as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$).

Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the effects of $\e(R)$ and $\var(R)$ are as we hypothesized (figure \@ref(fig:5-5-reg)). In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The relationships were estimated using a Generalized Linear Model with a Gamma family of distributions that accounted for the effects of both $\\e(R)$ and $\\var(R)$ as well as differences between the two quantiles. The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile. Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlinear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

@@

@@

Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021

* failing to account for $\var(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
* find examples of extreme events that caused population collapses or mass die-offs
* increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness

@@

@@

In the case of animals whose spatial needs change over their lifetimes [arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017], one may define a HR as a function of time with distinct HRs for each period or as a continuous function of time. In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available. However, when an animal's spatial needs change over time, models which assume range residency are not appropriate. Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results. Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].

@@

competition can promote niche partitioning [@fox_niche_1981]

Patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006].
